---
title: "PS5 Markdown Basic Plots Assignment"
author: "Shawna Nguyen"
date: "2023-02-18"
output: html_document
---

## Part 1

### Question 1.2

```{r, message=FALSE}
library(tidyverse)
world <- read_delim("gapminder.csv")
```

```{r}
world %>% 
  dim()
```

The *world* data contains `r nrow(world)` rows and `r ncol(world)` columns.

### Question 1.3 
```{r}
world %>% 
  sample_n(5) %>% 
  knitr::kable()
```

## Part 2

### Question 2.1
```{r}
world %>% 
  summarise(n = n_distinct(`iso3`))
```
There are 253 countries in the "iso3" dataset.

```{r}
world %>% 
  summarise(n = n_distinct(`iso2`))
```
There are 249 countries in the "iso2" dataset.

```{r}
world %>% 
  summarise(n = n_distinct(`name`))
```
There are 250 countries in the "name" dataset.

### Question 2.2.a
```{r}
world %>% 
  group_by(iso2) %>% 
  summarise(num = n_distinct(iso2)) %>% 
  arrange(desc(num))
```
There is exactly one iso-2 code that corresponds to every country.

### Question 2.2.b
```{r}
world %>% 
  group_by(name) %>% 
  summarise(num = n_distinct(iso3)) %>% 
  arrange(desc(num))
```
There is exactly one name and iso3-code that corresponds to every country.

```{r}
world %>% 
  select(iso3) %>% 
  filter(iso3 == c("CHANISL", "GBM", "KOS", "NLD_CURACAO")) %>% 
  group_by(iso3) %>% 
  sample_n(1)
```

The countries that have more than one iso-3 code are CHANISL, GBM, KOS and 
NLD CURACAO.

### Question 2.3
```{r}
world %>% 
  filter(!is.na(time)) %>% 
  summarise(min=min(time), max=max(time))
```

The latest year that this dataframe dates back to is 1960, while the most
recent year is 2019.

## Part 3

### Question 3.1
```{r}
world %>%
  group_by(time) %>% 
  filter(is.na(co2), is.na(co2_PC)) %>% 
  count() %>% 
  arrange(desc(n))
```

The years that have the most missing data is 2017, 2018 and 2019 with 217 NA's.

### Question 3.2
```{r}
world %>% 
  filter(!is.na(co2)) %>% 
  filter(name == "United States of America" | name == "China" | name == "India" | name == "Chile" | name == "Austria") %>% 
  ggplot(aes(time, co2, col=factor(name))) +
  geom_line() +
  labs(x = "Time", y = "CO2 Emissions", col = "Countries\n")
```

This line plot displays five different countries and their CO2 emissions over 
the years from 1960 to 2019. Looking at the graph, it shows me that China is the 
largest emitter of CO2 since 2016 with the United States coming in second. India
seems to be increasing their CO2 emissions over the years as well, but not as 
much as China and the US. On the other hand, Austria and Chile seems to emit
little to no CO2 compared to the other three countries on the graph.

### Question 3.3
(5pt) Now let's analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the same countries. What does this figure suggest?
```{r}
world %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(name == "United States of America" | name == "China" | name == "India" | name == "Chile" | name == "Austria") %>% 
  ggplot(aes(time, co2_PC, col=factor(name))) +
  geom_line() +
  labs(x = "Time", y = "CO2 Emissions (per capita)", col = "Countries\n")
```

After analyzing the line plot above, it tells me that the US average citizens
contribute to the most CO2 emissions per capita out of the other countries 
listed. Austria's, Chile's, China's and India's citizens do not emit as much
CO2 per capita on average as US citizens.

### Question 3.4
```{r}
world %>% 
  group_by(region) %>% 
  filter(!is.na(region)) %>%
  filter(time == 2016) %>% 
  summarise(new_avg = mean(co2_PC, na.rm = TRUE))
```

Looking at the average emissions per capita during the year 2016, the region 
that emitted the most was Europe with 6.64 CO2. The lowest emissions during the 
same year was Africa with 1.20 CO2 per capita. 

### Question 3.5
```{r}
average_pc <- world %>% 
  select(region, co2_PC, time) %>%
  filter(!is.na(region), time %in% c("1960", "2016")) %>% 
  group_by(region, time) %>% 
  summarise(avgCO2 = mean(co2_PC, na.rm = TRUE))

ggplot(average_pc, aes(region, avgCO2, fill=factor(time))) +
  geom_col(position = "dodge") +
  labs(title = "Average CO2 Emissions Per Capita Across The Continents",
           x = "Continents",
           y = "Average CO2 Per Capita",
           fill = "Time (Year)")
```

### Question 3.6
```{r}
co2_emitters <- world %>% 
  select(region, name, co2_PC, time) %>% 
  filter(time == "2016", !is.na(region), !is.na(co2_PC)) %>% 
  arrange(region, desc(co2_PC)) %>% 
  group_by(region)
```
```{r}
co2_emitters %>% 
  top_n(3, co2_PC) %>% 
  head(6)
```

## Part 4

### Question 4.1
```{r}
gdp_1960 <- world %>% 
  group_by(name) %>% 
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy)) %>% 
  select(region, GDP_PC, lifeExpectancy, totalPopulation, time) %>% 
  filter(time == "1960")

ggplot(gdp_1960, aes(GDP_PC, lifeExpectancy, size = totalPopulation/1000000, col = factor(region))) +
  geom_point(alpha=0.5) +
  labs(x = "GDP Per Capita",
       y = "Life Exepectancy in Years",
       title = "GDP Per Capita vs. Life Ecpectancy in 1960",
       color = "Continent",
       size = "Population in Millions")
```

Looking at the scatterplot above, Europe and America have the highest GDP per
capita which increases their life expectancy’s. This graph shows the
relationship between life expectancy and GDP per capita.

### Question 4.2
```{r}
gdp_2019 <- world %>% 
  group_by(name) %>% 
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy)) %>% 
  select(region, GDP_PC, lifeExpectancy, totalPopulation, time) %>% 
  filter(time == "2019")

ggplot(gdp_2019, aes(GDP_PC, lifeExpectancy, size = totalPopulation/1000000, col = factor(region))) +
  geom_point(alpha=0.5) +
  labs(x = "GDP Per Capita",
       y = "Life Exepectancy in Years",
       title = "GDP Per Capita vs. Life Ecpectancy in 2019",
       color = "Continent",
       size = "Population in Millions")
```

### Question 4.3 

Over the last 60 years, the world has increased its population, thus the life
expectancy has increased for each region. Similar to the GDP per capita, it has 
increased. When we compare these two plots together, it shows us that as the 
population increases, the GDP per capita and life expectancy increases. However,
the overall trend of these two plots remain somewhat similar.

### Question 4.4

```{r}
world %>% 
  group_by(region) %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  select(lifeExpectancy, time, region) %>% 
  filter(time == "1960") %>% 
  summarise(average_le_1960 = mean(lifeExpectancy))
```
```{r}
world %>% 
  group_by(region) %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  filter(region !="") %>%
  select(lifeExpectancy, time, region) %>% 
  filter(time == "2019") %>% 
  summarise(average_le_2019 = mean(lifeExpectancy))
```

After computing the average life expectancy for each continent, I compared the
numbers from 1960 and 2019, which show me that the life expectancy increased.
This fits with my findings on the figures that I made where over the last 60 
years, the world has increased its population, thus the life expectancy has 
increased for each region. 

### Question 4.5
```{r}
world %>% 
  select(region, lifeExpectancy, time) %>% 
  filter(!is.na(region), !is.na(lifeExpectancy), !is.na(time)) %>% 
  filter(time %in% c("1960", "2016")) %>% 
  group_by(region, time) %>% 
  summarise(average = mean(lifeExpectancy)) %>% 
  mutate(previous = lag(average), growth = average - previous)
```

Looking at the growth, the highest growing continent in terms of life expectancy 
from 1960 to 2019 is Asia and Africa. Whereas the lowest growing continent is 
Europe.

### Question 4.6
(6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!
```{r}
world %>% 
  filter(!is.na(GDP_PC)) %>% 
  filter(time %in% c("1960", "2019")) %>% 
  ggplot(aes(GDP_PC, fill = factor(time))) +
  geom_histogram(alpha = 0.5, position = "dodge", bins = 30) +
  scale_fill_manual(values = c("pink", "purple"), labels = c("1960", "2019")) +
  labs(x = "GDP Per Capita",
       y = "Count",
       title = "GDP Per Capita in 1960 and 2019")
```

### Question 4.7
```{r}
ranks <- world %>% 
  select(name, lifeExpectancy, time) %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time %in% c("1960", "2019")) %>% 
  arrange(time, desc(lifeExpectancy)) %>% 
  group_by(time) %>% 
  mutate(rankings = rank(desc(lifeExpectancy)))

ranks %>% 
  filter(name == "United States of America")
```

### Question 4.8
```{r}
ranks %>% 
  select(name, time, rankings) %>% 
  group_by(time) %>% 
  mutate(total_n = max(rankings)) %>% 
  filter(name == "United States of America") %>% 
  summarise(relativerank = rankings/total_n)
```

### Last Question

I spent about 16 hours on this assignment.


